---
title: "Vignettes"
author: "Kristian Peters"
date: "09/09/2022"
output: pdf_document
---

# CliMetabolomics workshop: Seasonal variations in the chemodiversity of two bryophytes

## 1. Preparations

First, load required libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(parallel)               # Detect number of cpu cores
library(foreach)                # For multicore parallel
library(doMC)                   # For multicore parallel
library(RColorBrewer)           # For colors
library(multtest)               # For diffreport
library(MSnbase)                # MS features
library(xcms)                   # Swiss army knife for metabolomics
library(CAMERA)                 # Metabolite Profile Annotation
library(Spectra)                # Spectra package needed for XCMS3
library(mixOmics)               # Statistics for various Omics
library(vegan)
library(multcomp)               # For Tukey test
library(Hmisc)                  # For correlation test
library(gplots)                 # For fancy heatmaps
library(lme4)                   # Linear Models with Random Effects
library(lmerTest)               # Create p-values for LMER
library(ape)                    # Phylogeny
library(pvclust)                # Phylogeny
library(dendextend)             # Phylogeny
library(cba)                    # Phylogeny
library(phangorn)               # Phylogeny
library(ontologyIndex)          # Reading obo ontology files
library(webchem)                # Converting InChI to InChIKey
library(circlize)               # For sunburst plot
library(plotrix)                # For sunburst plot
library(Boruta)                 # Random-Forest BORUTA
library(rpart)                  # Regression trees
library(caret)                  # Swiss-army knife for statistics
library(pROC)                   # Evaluation metrics
library(PRROC)                  # Evaluation metrics
library(multiROC)               # Evaluation metrics
```

Define some parameters

```{r setup, include=FALSE}
# Working directory
setwd("~/Desktop/Projekte/Habilitation/Mosses/iestimate_mtbls709/")

# Data directory
mzml_dir <- "~/Desktop/Projekte/Habilitation/Mosses/ms2-dda/data/raw/"

# MS1 variables
polarity <- "positive"
pol <- substr(x=polarity, start=1, stop=3)
ppm <- 35
peakwidth <- c(4,21)
prefilter <- c(5,50)
#fitgauss <- FALSE
#verbose.columns <- FALSE
snthresh <- 11
mzdiff <- 0.01
noise <- 0
integrate <- 1
minfrac <- 0.7
bwindow <- 0.25
minfracspan <- 0.2
ms1_intensity_cutoff <- 500	          # approx. 0.01% (= log 100)

# MS2 variables
mzabs <- 0.01                         # Absolute mass error (in seconds) used for merging MS/MS spectra
mzppm <- 5 #5                         # ppm error used for merging MS/MS spectra
rtabs <- 5 #20                        # Retention time error (in seconds) used for merging MS/MS spectra
max.rt.range <- 20                    # Permitted retention time window (in seconds) of grouped MS1 precursors
max.mz.range <- 0.01                  # Permitted m/z window of grouped MS1 precursors
min.rt <- 10                          # Minimum retention time for selected precursors
max.rt <- 1020                        # Maximum retention time for selected precursors
min.mz <- 50                          # Minimum m/z value for selected precursors
max.mz <- 1500                        # Maximum m/z value for selected precursors
msms.intensity.threshold <- 100       # Minimum intensity value for MS/MS peaks

# Classifier options
mzabs <- 0.01                        # Absolute mass error (in seconds) used for merging MS/MS spectra
mzppm <- 5 #35                       # ppm error used for merging MS/MS spectra
msms.intensity.threshold <- 100 #10  # Minimum intensity value for MS/MS peaks

CANOPUS <- TRUE                      # Use CANOPUS classification instead of MetFamily
SIRIUS_VERSION <- 5
```

Load RData object containing prepared data objects

```{r setup, include=FALSE}
load("r_data.rData")
```

Load iESTIMATE functions

```{r setup, include=FALSE}
source("_functions.r")
```

## 2. Detection of peaks at MS1 level

Load files and define colors

```{r, include=FALSE}
# XCMS optimization parameter
setMSnbaseFastLoad(TRUE)

# MS1 variables
polarity <- "positive"
pol <- substr(x=polarity, start=1, stop=3)

# General variables
mzml_files_pos <- NULL
mzml_names_pos <- NULL
mzml_times_pos <- NULL

# Load files
mzml_files_pos <- list.files(mzml_dir, pattern="*.mzML", recursive=T, full.names=T)
mzml_files_pos <- mzml_files_pos[grep(pol, mzml_files_pos, invert=FALSE)]
mzml_files_pos <- mzml_files_pos[grep("MM8", mzml_files_pos, invert=TRUE)]

# Let's focus on one species
mzml_files_pos <- mzml_files_pos[grep("Marpol", mzml_files_pos, invert=FALSE, perl=TRUE)]

# Basenames of files without path and without extension
mzml_names_pos <- gsub('(.*)\\..*', '\\1', gsub('( |-|,)', '.', basename(mzml_files_pos)))

# Create phenodata based on species
mzml_pheno_pos <- as.factor(sapply(strsplit(as.character(mzml_names_pos), "_"), function(x) {
	nam <- x[3];
	nam;
}))
mzml_pheno_pos <- data.frame(sample_name=mzml_pheno_pos, sample_group=mzml_pheno_pos)
mzml_pheno_samples_pos <- as.factor(mzml_pheno_pos$sample_group)
mzml_pheno_colors_pos <- c("firebrick3", "chartreuse3", "darkgoldenrod3", "deepskyblue3")
mzml_pheno_colors_samples_pos <- sapply(mzml_pheno_samples_pos, function(x) { x <- mzml_pheno_colors_pos[which(x==levels(mzml_pheno_samples_pos))] } )
```

Extract chromatograms

```{r, include=FALSE}
# Import raw data as MSnbase object
raw_data_pos <- readMSData(files=mzml_files_pos, pdata=new("NAnnotatedDataFrame", mzml_pheno_pos), mode="onDisk", centroided=TRUE)

# Restrict data to max.rt seconds
raw_data_pos <- filterRt(raw_data_pos, c(0, max.rt))

# Inspect mz values per file
raw_mz_pos <- mz(raw_data_pos)
raw_mz_pos <- split(raw_mz_pos, f = fromFile(raw_data_pos))
print(length(raw_mz_pos))

# Get base peak chromatograms
chromas_pos <- chromatogram(raw_data_pos, aggregationFun="max")
```

Inspect chromatograms

```{r, include=FALSE}
# Plot chromatograms based on phenodata groups
par(mfrow=c(1,1), mar=c(4,4,4,1), oma=c(0,0,0,0), cex.axis=0.9, cex=0.6)
plot(chromas_pos, main="Raw chromatograms", xlab="retention time [s]", ylab="intensity", col=mzml_pheno_colors_samples_pos)
legend("topleft", bty="n", pt.cex=0.5, cex=0.7, y.intersp=0.7, text.width=0.5, pch=20, col=mzml_pheno_colors_pos, legend=mzml_pheno_samples_pos)
```

Plot TICs

```{r, include=FALSE}
# Get TICs
par(mfrow=c(1,1), mar=c(5,4,4,1), oma=c(0,0,0,0), cex.axis=0.9, cex=0.6)
tics_pos <- split(tic(raw_data_pos), f=fromFile(raw_data_pos))
boxplot(tics_pos, names=mzml_names_pos, las=2, col=mzml_pheno_colors_pos, ylab="intensity", main="Total ion current")
```

Perform binning of samples

```{r, include=FALSE}
# Grouping/binning the samples based on similarity of their base peak chromatogram to spot potentially problematic samples
chromas_bin_pos <- MSnbase::bin(chromas_pos, binSize=2)
chromas_bin_cor_pos <- cor(log2(do.call(cbind, lapply(chromas_bin_pos, intensity))))
colnames(chromas_bin_cor_pos) <- rownames(chromas_bin_cor_pos) <- raw_data_pos$sample_name
par(mfrow=c(1,1), mar=c(4,4,4,1), oma=c(0,0,0,0), cex.axis=0.9, cex=0.6)
heatmap(chromas_bin_cor_pos)
```

Detection of chromatographic peaks

```{r, include=FALSE}
# Peak detection in MS1 data
ms1_params_pos <- CentWaveParam(ppm=ppm, mzCenterFun="mean", peakwidth=peakwidth, prefilter=prefilter, mzdiff=mzdiff, snthresh=snthresh, noise=noise, integrate=integrate, firstBaselineCheck=TRUE, verboseColumns=FALSE, fitgauss=FALSE, roiList=list(), roiScales=numeric())
ms1_data_pos <- findChromPeaks(raw_data_pos, param=ms1_params_pos)
```

Plot short summary

```{r, include=FALSE}
# Per file summary
ms1_summary_pos <- lapply(split.data.frame(chromPeaks(ms1_data_pos), f=chromPeaks(ms1_data_pos)[, "sample"]), FUN=function(z) { c(peak_count=nrow(z), rt=quantile(z[, "rtmax"] - z[, "rtmin"])) } )
ms1_summary_pos <- do.call(rbind, ms1_summary_pos)
rownames(ms1_summary_pos) <- basename(fileNames(ms1_data_pos))
print(ms1_summary_pos)
table(msLevel(ms1_data_pos))
```

Plot overview

```{r, include=FALSE}
# To get a global overview of the peak detection we can plot the frequency of identified peaks per file along the retention time axis. This allows to identify time periods along the MS run in which a higher number of peaks was identified and evaluate whether this is consistent across files.
par(mfrow=c(1,1), mar=c(4,18,4,1), oma=c(0,0,0,0), cex.axis=0.9, cex=0.6)
plotChromPeakImage(ms1_data_pos, main="Frequency of identified peaks per RT")
```

Group peaks

```{r, include=FALSE}
# Group peaks
ms1_data_pos <- groupChromPeaks(ms1_data_pos, param=PeakDensityParam(sampleGroups=ms1_data_pos$sample_group, minFraction=minfrac, bw=bwindow))
```

Retention time correction

```{r, include=FALSE}
# RT correction
ms1_data_pos <- adjustRtime(ms1_data_pos, param=PeakGroupsParam(minFraction=minfrac,smooth="loess",span=minfracspan,family="gaussian"))

# Plot the difference of raw and adjusted retention times
par(mfrow=c(2,1), mar=c(4.5,4.2,4,1), cex=0.8)
plot(chromas_pos, peakType="none", main="Raw chromatograms", col=mzml_pheno_colors_samples_pos)
plotAdjustedRtime(ms1_data_pos, lwd=2, main="Retention Time correction", col=mzml_pheno_colors_samples_pos)
par(mfrow=c(1,1), mar=c(4,4,4,1), oma=c(0,0,0,0), cex.axis=0.9, cex=0.8)
```

Re-group peaks

```{r, include=FALSE}
# Group peaks
ms1_data_pos <- groupChromPeaks(ms1_data_pos, param=PeakDensityParam(sampleGroups=ms1_data_pos$sample_group, minFraction=minfrac, bw=bwindow))

# Get integrated peak intensity per feature/sample
print(head(featureValues(ms1_data_pos, value="into")))
```

Peak filling

```{r, include=FALSE}
# Fill peaks
#ms1_data_pos <- fillChromPeaks(ms1_data_pos, param=FillChromPeaksParam(ppm=ppm, fixedRt=0, expandRt=5))
#head(featureValues(ms1_data_pos))
#head(featureSummary(ms1_data_pos, group=ms1_data_pos$sample_group))
```

Inspect the data objects one more time

```{r, include=FALSE}
# Show peaks
tail(chromPeaks(ms1_data_pos))
tail(chromPeakData(ms1_data_pos))

# Show process history
processHistory(ms1_data_pos)
```

Build MS1 feature tables

```{r, include=FALSE}
# Build feature matrix
ms1_matrix_pos <- featureValues(ms1_data_pos, method="medret", value="into")
colnames(ms1_matrix_pos) <- mzml_names_pos
dim(ms1_matrix_pos)
feat_list_pos <- t(ms1_matrix_pos)

# Build feature summary
ms1_summary_pos <- featureSummary(ms1_data_pos)
ms1_def_pos <- featureDefinitions(ms1_data_pos)

# Transform data
feat_list_pos <- log(feat_list_pos)

# Missing value imputation
feat_list_pos[is.na(feat_list_pos)] <- 0
feat_list_pos[which(feat_list_pos < 0)] <- 0
feat_list_pos[is.infinite(feat_list_pos)] <- 0
feat_list_pos <- feat_list_pos[!apply(feat_list_pos, MARGIN=1, function(x) max(x,na.rm=TRUE) == min(x,na.rm=TRUE)),]

# Create single 0/1 matrix
bina_list_pos <- feat_list_pos
bina_list_pos[bina_list_pos < log(ms1_intensity_cutoff)] <- 0
bina_list_pos[bina_list_pos != 0] <- 1

# Only unique compounds in group mzml_pheno$ and not the others
uniq_list_pos <- apply(X=bina_list_pos, MARGIN=2, FUN=function(x) { if (length(unique(mzml_pheno_pos$sample_group[grepl("1", x)])) == 1) x else rep(0, length(x)) } )
colnames(uniq_list_pos) <- colnames(bina_list_pos)
rownames(uniq_list_pos) <- rownames(bina_list_pos)

# Create matrix with diversity measures
model_div_pos             <- data.frame(features=apply(X=bina_list_pos, MARGIN=1, FUN=function(x) { sum(x) } ))
model_div_pos$richness    <- apply(X=bina_list_pos, MARGIN=1, FUN=function(x) { sum(x) } )
#model_div_pos$menhinick   <- apply(X=bina_list_pos, MARGIN=1, FUN=function(x) { menhinick.diversity(x) } )
model_div_pos$shannon     <- apply(X=feat_list_pos, MARGIN=1, FUN=function(x) { vegan::diversity(x, index="shannon") })
model_div_pos$pielou      <- apply(X=scale(feat_list_pos, center=F), MARGIN=1, FUN=function(x) { vegan::diversity(x, index="shannon") / log(vegan::specnumber(x)) })
#model_div_pos$chao        <- vegan::specpool2vect(X=vegan::specpool(feat_list_pos, species), index="chao")
model_div_pos$simpson     <- apply(X=feat_list_pos, MARGIN=1, FUN=function(x) { vegan::diversity(x, index="simpson") })
model_div_pos$inverse     <- apply(X=feat_list_pos, MARGIN=1, FUN=function(x) { vegan::diversity(x, index="inv") })
model_div_pos$fisher      <- apply(X=feat_list_pos, MARGIN=1, FUN=function(x) { fisher.alpha(round(x,0)) })
model_div_pos$unique      <- apply(X=uniq_list_pos, MARGIN=1, FUN=function(x) { sum(x) })

# Remove NAs if present
model_div_pos[is.na(model_div_pos)] <- 0
```

## 3. Statistics at MS1 level

### 3.1. Histogram

```{r, include=FALSE}
hist(as.numeric(feat_list_pos))
```

### 3.2. PCA

```{r, include=FALSE}
model_pca_pos <- prcomp(feat_list_pos)

plot(model_pca_pos$x[, 1], model_pca_pos$x[,2], pch=19, main="PCA",
	 xlab=paste0("PC1: ", format(summary(model_pca_pos)$importance[2, 2] * 100, digits=3), " % variance"),
	 ylab=paste0("PC2: ", format(summary(model_pca_pos)$importance[2, 3] * 100, digits=3), " % variance"),
	 col=mzml_pheno_colors_samples_pos, bg=mzml_pheno_colors_samples_pos, cex=2)
grid()
text(model_pca_pos$x[, 1], model_pca_pos$x[,2], labels=mzml_names_pos, col=mzml_pheno_colors_samples_pos, pos=3, cex=0.5)
```

### 3.3 Diversity measures

Chemical richness

```{r, include=FALSE}
boxplot(model_div_pos$features ~ mzml_pheno_samples_pos, col=mzml_pheno_colors_pos, names=NA, main="Number of features", xlab="treatment", ylab="number of features")
text(1:length(unique(mzml_pheno_samples_pos)), par("usr")[3]-(par("usr")[4]-par("usr")[3])/14, srt=-22.5, adj=0.5, labels=levels(mzml_pheno_samples_pos), xpd=TRUE, cex=0.9)
div_tukey <- tukey.test(response=model_div_pos$features, term=as.factor(mzml_pheno_samples_pos))
text(1:length(unique(mzml_pheno_samples_pos)), par("usr")[4]+(par("usr")[4]-par("usr")[3])/40, adj=0.5, labels=div_tukey[,1], xpd=TRUE, cex=0.8)
```

Shannon diversity index

```{r, include=FALSE}
boxplot(model_div_pos$shannon ~ mzml_pheno_samples_pos, col=mzml_pheno_colors_pos, names=NA, main="Shannon diversity (H\')", xlab="treatment", ylab="Shannon diversity index (H\')")
text(1:length(unique(mzml_pheno_samples_pos)), par("usr")[3]-(par("usr")[4]-par("usr")[3])/14, srt=-22.5, adj=0.5, labels=levels(mzml_pheno_samples_pos), xpd=TRUE, cex=0.9)
div_tukey <- tukey.test(response=model_div_pos$shannon, term=as.factor(mzml_pheno_samples_pos))
text(1:length(unique(mzml_pheno_samples_pos)), par("usr")[4]+(par("usr")[4]-par("usr")[3])/40, adj=0.5, labels=div_tukey[,1], xpd=TRUE, cex=0.8)
```

Pielou's evenness index

```{r, include=FALSE}
boxplot(model_div_pos$pielou ~ mzml_pheno_samples_pos, col=mzml_pheno_colors_pos, names=NA, main="Pielou\'s evenness", xlab="treatment", ylab="Pielou diversity index (J)")
text(1:length(unique(mzml_pheno_samples_pos)), par("usr")[3]-(par("usr")[4]-par("usr")[3])/14, srt=-22.5, adj=0.5, labels=levels(mzml_pheno_samples_pos), xpd=TRUE, cex=0.9)
div_tukey <- tukey.test(response=model_div_pos$pielou, term=as.factor(mzml_pheno_samples_pos))
text(1:length(unique(mzml_pheno_samples_pos)), par("usr")[4]+(par("usr")[4]-par("usr")[3])/40, adj=0.5, labels=div_tukey[,1], xpd=TRUE, cex=0.8)
```

### 3.4. Variable Selection

Selection of features

```{r, include=FALSE}
# Random Forest
sel_rf_pos <- f.select_features_random_forest(feat_matrix=feat_list_pos, sel_factor=as.factor(mzml_pheno_samples_pos), sel_colors=mzml_pheno_colors_pos, tune_length=10, quantile_threshold=0.90, plot_roc_filename="plots/pos_ms1_select_rf_roc.pdf")
print(paste("Number of selected features:", f.count.selected_features(sel_feat=sel_rf_pos$`_selected_variables_`)))
f.heatmap.selected_features(feat_list=feat_list_pos, sel_feat=sel_rf_pos$`_selected_variables_`, sel_names=paste0("     ",sel_rf_pos$`_selected_variables_`), filename="plots/pos_ms1_select_rf.pdf", main="Random Forest", plot_width=6, plot_height=5, cex_col=0.5, cex_row=0.4)
```

```{r, include=FALSE}
# PLS
sel_pls_pos <- f.select_features_pls(feat_matrix=feat_list_pos, sel_factor=as.factor(mzml_pheno_samples_pos), sel_colors=mzml_pheno_colors_pos, components=length(unique(mzml_pheno_samples_pos))-1, tune_length=10, quantile_threshold=0.982, plot_roc_filename="plots/pos_ms1_select_pls_roc.pdf")
print(paste("Number of selected features:", f.count.selected_features(sel_feat=sel_pls_pos$`_selected_variables_`)))
f.heatmap.selected_features(feat_list=feat_list_pos, sel_feat=sel_pls_pos$`_selected_variables_`, sel_names=paste0("     ",sel_pls_pos$`_selected_variables_`), filename="plots/pos_ms1_select_pls.pdf", main="PLS", plot_width=6, plot_height=5, cex_col=0.5, cex_row=0.4)
```

## 4. Detection of fragment spectra at MS2 level

Detect fragment spectra and correlating them to the MS1 precursors:

```{r, include=FALSE}
# Estimate precursor intensity
precursor_intensity_pos <- estimatePrecursorIntensity(ms1_data_pos)
print(head(na.omit(precursor_intensity_pos)))

# Reconstruct MS2 spectra from MS1 data
ms2_data_pos <- chromPeakSpectra(ms1_data_pos, msLevel=2L, return.type="Spectra")
print(ms2_data_pos)
print(length(ms2_data_pos$peak_id))

# Extract all usable MS2 spectra
ms2_spectra_pos <- list()
#for (i in 1:nrow(ms1_def_pos)) {
ms2_spectra_pos <- foreach(i=1:nrow(ms1_def_pos)) %dopar% {
	# Extract existing MS2 spectra for feature
	feature_of_interest <- ms1_def_pos[i, "mzmed"]
	peaks_of_interest <- chromPeaks(ms1_data_pos, mz=feature_of_interest, ppm=ppm)
	
	# Continue if feature has MS2 peaks
	if (length(which(ms2_data_pos$peak_id %in% rownames(peaks_of_interest))) > 0) {
		# Extract spectra
		spectra_of_interest <- ms2_data_pos[ms2_data_pos$peak_id %in% rownames(peaks_of_interest)]
		combined_spectra_of_interest <- filterIntensity(spectra_of_interest, intensity=c(1,Inf), backend=MsBackendDataFrame)
		combined_spectra_of_interest <- setBackend(combined_spectra_of_interest, backend=MsBackendDataFrame())
		
		# Combine spectra
		combined_spectra_of_interest <- Spectra::combineSpectra(combined_spectra_of_interest, FUN=combinePeaks, ppm=ppm, peaks="union", minProp=0.8, intensityFun=median, mzFun=median, backend=MsBackendDataFrame)#f=rownames(peaks_of_interest))
		
		# Remove noise from spectra
		#combined_spectra_of_interest <- pickPeaks(combined_spectra_of_interest, snr=1.0, method="SuperSmoother") #MAD
		#combined_spectra_of_interest <- Spectra::smooth(combined_spectra_of_interest, method="SavitzkyGolay") #(Weighted)MovingAverage
		
		# Only keep spectral data
		combined_spectra_peaks <- as.data.frame(peaksData(combined_spectra_of_interest)[[1]])
		
		# Plot merged spectrum
		#Spectra::plotSpectra(combined_spectra_of_interest)
		#plot(x=combined_spectra_peaks[,1], y=combined_spectra_peaks[,2], type="h", xlab="m/z", ylab="intensity", main=paste("Precursor m/z",combined_spectra_of_interest@backend@spectraData$precursorMz[[1]]))
		#length(spectra_of_interest$peak_id)
		
		#ms2_spectra_pos[[i]] <- combined_spectra_of_interest
		return(combined_spectra_of_interest)
	}
}

# Remove empty spectra
names(ms2_spectra_pos) <- rownames(ms1_def_pos)
ms2_spectra_pos <- ms2_spectra_pos[lengths(ms2_spectra_pos) != 0]

# Relate all MS2 spectra to MS1 precursors
ms1_def_pos$has_ms2 <- as.integer(rownames(ms1_def_pos) %in% names(ms2_spectra_pos))
print(paste0("Number of MS2 spectra related to precursor: ", length(which(ms1_def_pos$has_ms2>0))))
```

Save spectra in MGF format for processing with SIRIUS

```{r, include=FALSE}
# Save all MS2 spectra in MGF file
mgf_text <- NULL
for (i in names(ms2_spectra_pos)) {
	mgf_text <- c(mgf_text, paste0("COM=", i))
	mgf_text <- c(mgf_text, "BEGIN IONS")
	mgf_text <- c(mgf_text, "MSLEVEL=2")
	mgf_text <- c(mgf_text, paste0("TITLE=", i))
	mgf_text <- c(mgf_text, paste0("RTINSECONDS=", ms1_def_pos[i, "rtmed"]))
	mgf_text <- c(mgf_text, paste0("PEPMASS=", ms1_def_pos[i, "mzmed"]))
	if (polarity == "positive") {
		mgf_text <- c(mgf_text, paste0("CHARGE=", "1+"))
	} else {
		mgf_text <- c(mgf_text, paste0("CHARGE=", "1-"))
	}
	mgf_text <- c(mgf_text, paste(as.data.frame(peaksData(ms2_spectra_pos[[i]])[[1]])$mz, as.data.frame(peaksData(ms2_spectra_pos[[i]])[[1]])$intensity, sep=" "))
	mgf_text <- c(mgf_text, "END IONS")
	mgf_text <- c(mgf_text, "")
}

# Write MGF file
cat(mgf_text, file="data/pos_ms2_spectra.mgf", sep="\n")
```

-----\> SIRIUS / CANOPUS \<-----

## 5. Processing CANOPUS classification data

Build data objects

```{r, include=FALSE}
# Apply classified classes onto feature table
ms1_def_pos$primary_class <- ""
ms1_def_pos$alternative_classes <- ""

# Read SIRIUS/CANOPUS classifier
classifier_canopus_pos <- read.table(file=paste0("data/pos_ms2_canopus_summary.tsv"), header=TRUE, sep="\t", quote="\"", fill=FALSE, dec=".", stringsAsFactors=FALSE)#, encoding="UTF-8", fileEncoding="UTF-8")
if (SIRIUS_VERSION == 4) {
	classifier_canopus_pos$Metabolite.name <- gsub(x=classifier_canopus_pos$name, pattern=".*_", replacement="")
	classifier_canopus_pos$primary_class <- paste("Organic compounds", classifier_canopus_pos$superclass, classifier_canopus_pos$class, classifier_canopus_pos$subclass, classifier_canopus_pos$level.5, sep="; ")
	classifier_canopus_pos$primary_class <- gsub(x=classifier_canopus_pos$primary_class, pattern="; ; ", replacement="; ", perl=TRUE)
	classifier_canopus_pos$primary_class <- gsub(x=classifier_canopus_pos$primary_class, pattern="; ; ", replacement="; ", perl=TRUE)
	classifier_canopus_pos$primary_class <- gsub(x=classifier_canopus_pos$primary_class, pattern="; $", replacement="", perl=TRUE)
	classifier_canopus_pos$primary_class <- gsub(x=classifier_canopus_pos$primary_class, pattern="(\\'|\\>|\\(|\\))", replacement="", perl=TRUE)
	classifier_canopus_pos$Annotation..putative. <- classifier_canopus_pos$primary_class
	classifier_canopus_pos$alternative_classes <- classifier_canopus_pos$all.classifications
} else {
	sirius_identification_pos <- read.table(file=paste0("data/pos_ms2_compound_identifications.tsv"), header=TRUE, sep="\t", quote="\"", fill=FALSE, dec=".", stringsAsFactors=FALSE)#, encoding="UTF-8", fileEncoding="UTF-8")
	classifier_canopus_pos$Metabolite.name <- ""
	for (i in 1:length(classifier_canopus_pos$id)) {
		x = which(gsub(x=sirius_identification_pos$id, pattern=".*?_", replacement="", perl=TRUE) %in% gsub(x=classifier_canopus_pos$id[i], pattern=".*?_", replacement="", perl=TRUE))
		if (length(x) > 0) classifier_canopus_pos$Metabolite.name[i] <- gsub(x=sirius_identification_pos$id[x], pattern=".*?_", replacement="", perl=TRUE)
	}
	classifier_canopus_pos$Metabolite.name[classifier_canopus_pos$Metabolite.name == "null"] <- ""
	classifier_canopus_pos$primary_class <- paste("Organic compounds", classifier_canopus_pos$ClassyFire_superclass, classifier_canopus_pos$ClassyFire_class, classifier_canopus_pos$ClassyFire_subclass, classifier_canopus_pos$ClassyFire_level.5, sep="; ")
	classifier_canopus_pos$primary_class <- gsub(x=classifier_canopus_pos$primary_class, pattern="; ; ", replacement="; ", perl=TRUE)
	classifier_canopus_pos$primary_class <- gsub(x=classifier_canopus_pos$primary_class, pattern="; ; ", replacement="; ", perl=TRUE)
	classifier_canopus_pos$primary_class <- gsub(x=classifier_canopus_pos$primary_class, pattern="; $", replacement="", perl=TRUE)
	classifier_canopus_pos$primary_class <- gsub(x=classifier_canopus_pos$primary_class, pattern="(\\'|\\>|\\(|\\))", replacement="", perl=TRUE)
	classifier_canopus_pos$Annotation..putative. <- classifier_canopus_pos$primary_class
	classifier_canopus_pos$alternative_classes <- classifier_canopus_pos$all.classifications
}

for (j in unique(classifier_canopus_pos$Metabolite.name)) {
	obj <- classifier_canopus_pos[classifier_canopus_pos$Metabolite.name %in% j, "Annotation..putative."]
	ms1_def_pos[j, "primary_class"] <- obj[1]
}

# Create CANOPUS classifier object for each sample
classifiers_pos <- list()
for (i in mzml_names_pos) {
	obj <- names(which(bina_list_pos[rownames(bina_list_pos)==i, colnames(bina_list_pos) %in% classifier_canopus_pos$Metabolite.name] > 0))
	classifiers_pos[[i]] <- classifier_canopus_pos[classifier_canopus_pos$Metabolite.name %in% obj, ]
}

# Diversity of classes per sample
div_classes_samples_pos <- NULL
for (i in mzml_names_pos) {
	obj <- table(classifiers_pos[[i]][,"Annotation..putative."])
	obj <- data.frame(classes=names(obj), frequency=as.numeric(obj))
	if (is.null(div_classes_samples_pos)) {
		div_classes_samples_pos <- obj
	} else {
		div_classes_samples_pos <- merge(div_classes_samples_pos, obj, by="classes", all.x=TRUE, all.y=TRUE)
	}
}
rownames(div_classes_samples_pos) <- div_classes_samples_pos$classes
div_classes_samples_pos <- div_classes_samples_pos[, -which(colnames(div_classes_samples_pos)=="classes")]
colnames(div_classes_samples_pos) <- mzml_names_pos

# Diversity of classes
div_classes_pos <- div_classes_samples_pos
div_classes_pos[is.na(div_classes_pos)] <- 0
div_classes_pos <- apply(X=div_classes_pos, MARGIN=1, FUN=function(x) { sum(x) })
div_classes_pos <- data.frame(row.names=names(div_classes_pos), frequency=as.numeric(div_classes_pos))

# Plot diversity of classes in all samples
par(mfrow=c(1,1), mar=c(4,15,4,1), oma=c(0,0,0,0), cex.axis=0.8, cex=0.9)
barplot(div_classes_pos$frequency, names.arg=gsub('.*; ','',rownames(div_classes_pos)), las=1, horiz=TRUE, xlab="frequency", main="Diversity of compound classes", col=rainbow(n=nrow(div_classes_pos), alpha=0.6))

# Sunburst plot of classes of all samples
par(mfrow=c(1,1), mar=c(0,0,0,0), oma=c(0,0,0,0), cex.axis=1, cex=1)
sunBurstPlotFromSubstanceClasses(rownames(div_classes_pos), div_classes_pos$frequency, colorStart=0.0, colorAlpha=0.6)
write.csv(data.frame("Compound classes"=rownames(div_classes_pos), "Number of spectra"=div_classes_pos$frequency), file="plots/pos_ms2_classes_sunburst.csv", row.names=FALSE)

# Classes
classes_pos <- rownames(div_classes_pos)
classes_pos <- classes_pos[which(grepl(pattern="^Organic compounds", x=classes_pos))]
classes_pos <- gsub(x=classes_pos, pattern=":", replacement="")
classes_pos <- gsub(x=classes_pos, pattern="/", replacement="; ")

# Imputation of NA with zeros
div_classes_pos[is.na(div_classes_pos)] <- 0
div_classes_samples_pos[is.na(div_classes_samples_pos)] <- 0

# Classification list for statistics
class_list_pos <- as.data.frame(t(div_classes_samples_pos))
class_list_pos[is.na(class_list_pos)] <- 0

# Log Transform
#class_list_pos <- log2(class_list_pos + 1)

# Only keep class names
colnames(class_list_pos) <- gsub(x=colnames(class_list_pos), pattern='.*; ', replacement='')

# Generate class_int_list_pos with abundances instead of counts
class_int_list_pos <- class_list_pos

for (i in 1:nrow(class_list_pos)) {
	samp <- rownames(class_list_pos)[i]
	for (j in 1:ncol(class_list_pos)) {
		cl <- colnames(class_list_pos)[j]
		ft <- classifier_canopus_pos$Metabolite.name[which(gsub(x=classifier_canopus_pos$primary_class, pattern='.*; ', replacement='') == cl)]
		ints <- as.numeric(feat_list_pos[i, which(colnames(feat_list_pos) %in% ft)])
		class_int_list_pos[i, j] <- sum(ints) * as.numeric(class_list_pos[i, j])
	}
}
```

Classification can be performed at different levels of the ontology

```{r, include=FALSE}
# Make superclasses at level 3
superclass_level_pos <- 3
div_superclasses_samples_names_pos <- NULL
for (i in c(1:superclass_level_pos)) {
	div_superclasses_samples_names_pos <- c(div_superclasses_samples_names_pos, lapply(X=strsplit(rownames(div_classes_samples_pos), '; '), FUN=function(x) { gsub(x=paste(x[1:i],sep='',collapse='; '),pattern='; NA',replacement='') }))
}
div_superclasses_samples_pos <- data.frame()
for (i in c(1:ncol(div_classes_samples_pos))) div_superclasses_samples_pos <- rbind(div_superclasses_samples_pos, rep(0, length(unique(div_superclasses_samples_names_pos))))
div_superclasses_samples_pos <- t(div_superclasses_samples_pos)
colnames(div_superclasses_samples_pos) <- colnames(div_classes_samples_pos)
rownames(div_superclasses_samples_pos) <- unique(div_superclasses_samples_names_pos)
for (i in rownames(div_superclasses_samples_pos)) {
	for (j in c(1:ncol(div_classes_samples_pos))) {
		div_superclasses_samples_pos[rownames(div_superclasses_samples_pos)==i, j] <- sum(div_classes_samples_pos[grep(x=rownames(div_classes_samples_pos), pattern=i), j])
	}
}

# Diversity of superclasses
div_superclasses_pos <- div_superclasses_samples_pos
div_superclasses_pos[is.na(div_superclasses_pos)] <- 0
div_superclasses_pos <- apply(X=div_superclasses_pos, MARGIN=1, FUN=function(x) { sum(x) })
div_superclasses_pos <- data.frame(row.names=names(div_superclasses_pos), frequency=as.numeric(div_superclasses_pos))

# Sunburst plot
sunBurstPlotFromSubstanceClasses(rownames(div_superclasses_pos), div_superclasses_pos$frequency, colorStart=0.0, colorAlpha=0.6)
write.csv(data.frame("Compound classes"=rownames(div_superclasses_pos), "Number of spectra"=div_superclasses_pos$frequency), file="plots/pos_ms2_superclasses_sunburst.csv", row.names=FALSE)

# Imputation of NA with zeros
div_superclasses_pos[is.na(div_superclasses_pos)] <- 0
div_superclasses_samples_pos[is.na(div_superclasses_samples_pos)] <- 0

# Classification list for statistics
superclass_list_pos <- as.data.frame(t(div_superclasses_samples_pos))
superclass_list_pos[is.na(superclass_list_pos)] <- 0

# Log Transform
#superclass_list_pos <- log2(superclass_list_pos + 1)

# Only keep superclass names
colnames(superclass_list_pos) <- gsub(x=colnames(superclass_list_pos), pattern='.*; ', replacement='')

# Generate superclass_int_list_pos with abundances instead of counts
superclass_int_list_pos <- superclass_list_pos

for (i in 1:nrow(superclass_list_pos)) {
	samp <- rownames(superclass_list_pos)[i]
	for (j in 1:ncol(superclass_list_pos)) {
		cl <- colnames(superclass_list_pos)[j]
		ft <- classifier_canopus_pos$Metabolite.name[which(classifier_canopus_pos$primary_class %in% classifier_canopus_pos$primary_class[grep(x=classifier_canopus_pos$primary_class, pattern=cl)]) ]
		ints <- as.numeric(feat_list_pos[i, which(colnames(feat_list_pos) %in% ft)])
		superclass_int_list_pos[i, j] <- sum(ints) * as.numeric(superclass_list_pos[i, j])
	}
}
```

## 6. Statistics at MS2 level

### 6.1. Histogram

```{r, include=FALSE}
hist(as.numeric(unlist(class_list_pos)), main="Histogram", xlab="class_list")
```

### 6.2. PCA

```{r, include=FALSE}
model_pca_pos <- prcomp(class_list_pos)

plot(model_pca_pos$x[, 1], model_pca_pos$x[, 2], pch=19, main="PCA",
	 xlab=paste0("PC2: ", format(summary(model_pca_pos)$importance[2, 1] * 100, digits=3), " % variance"),
	 ylab=paste0("PC3: ", format(summary(model_pca_pos)$importance[2, 2] * 100, digits=3), " % variance"),
	 col=mzml_pheno_colors_samples_pos, bg=mzml_pheno_colors_samples_pos, cex=2)
grid()
text(model_pca_pos$x[, 1], model_pca_pos$x[, 2], labels=mzml_names_pos, col=mzml_pheno_colors_samples_pos, pos=3, cex=0.5)
```

### 6.3. Variable selection

```{r, include=FALSE}
# Random Forest
sel_rf_pos <- f.select_features_random_forest(feat_matrix=class_list_pos, sel_factor=mzml_pheno_samples_pos, sel_colors=mzml_pheno_colors_pos, tune_length=10, quantile_threshold=0.95, plot_roc_filename="plots/pos_ms2_classes_select_rf_roc.pdf")
print(paste("Number of selected features:", f.count.selected_features(sel_feat=sel_rf_pos$`_selected_variables_`)))
f.heatmap.selected_features(feat_list=class_list_pos, sel_feat=sel_rf_pos$`_selected_variables_`, sel_names=paste0("     ",sel_rf_pos$`_selected_variables_`), filename="plots/pos_ms2_classes_select_rf.pdf", main="Random Forest", plot_width=6, plot_height=5, cex_col=0.5, cex_row=0.4)

# PLS
sel_pls_pos <- f.select_features_pls(feat_matrix=class_list_pos, sel_factor=mzml_pheno_samples_pos, sel_colors=mzml_pheno_colors_pos, components=length(unique(mzml_pheno_samples_pos))-1, tune_length=10, quantile_threshold=0.9, plot_roc_filename="plots/pos_ms2_classes_select_pls_roc.pdf")
print(paste("Number of selected features:", f.count.selected_features(sel_feat=sel_pls_pos$`_selected_variables_`)))
f.heatmap.selected_features(feat_list=class_list_pos, sel_feat=sel_pls_pos$`_selected_variables_`, sel_names=paste0("     ",sel_pls_pos$`_selected_variables_`), filename="plots/pos_ms2_classes_select_pls.pdf", main="PLS", plot_width=6, plot_height=5, cex_col=0.5, cex_row=0.4)
```

## 7. Statistics at MS2 level (Superclass level)

### 7.1. PCA

```{r, include=FALSE}
model_pca_pos <- prcomp(superclass_list_pos)

plot(model_pca_pos$x[, 1], model_pca_pos$x[, 2], pch=19, main="PCA",
	 xlab=paste0("PC2: ", format(summary(model_pca_pos)$importance[2, 1] * 100, digits=3), " % variance"),
	 ylab=paste0("PC3: ", format(summary(model_pca_pos)$importance[2, 2] * 100, digits=3), " % variance"),
	 col=mzml_pheno_colors_samples_pos, bg=mzml_pheno_colors_samples_pos, cex=2)
grid()
text(model_pca_pos$x[, 1], model_pca_pos$x[, 2], labels=mzml_names_pos, col=mzml_pheno_colors_samples_pos, pos=3, cex=0.5)
```

### 7.2. Variable selection

```{r, include=FALSE}
# Random Forest
sel_rf_pos <- f.select_features_random_forest(feat_matrix=superclass_list_pos, sel_factor=mzml_pheno_samples_pos, sel_colors=mzml_pheno_colors_pos, tune_length=10, quantile_threshold=0.95, plot_roc_filename="plots/pos_ms2_superclasses_select_rf_roc.pdf")
print(paste("Number of selected features:", f.count.selected_features(sel_feat=sel_rf_pos$`_selected_variables_`)))
f.heatmap.selected_features(feat_list=superclass_list_pos, sel_feat=sel_rf_pos$`_selected_variables_`, filename="plots/pos_ms2_superclasses_select_rf.pdf", main="Random Forest", plot_width=6, plot_height=5, cex_col=0.5, cex_row=0.4)

# PLS
sel_pls_pos <- f.select_features_pls(feat_matrix=superclass_list_pos, sel_factor=mzml_pheno_samples_pos, sel_colors=mzml_pheno_colors_pos, components=length(unique(mzml_pheno_samples_pos))-1, tune_length=10, quantile_threshold=0.95, plot_roc_filename="plots/pos_ms2_superclasses_select_pls_roc.pdf")
print(paste("Number of selected features:", f.count.selected_features(sel_feat=sel_pls_pos$`_selected_variables_`)))
f.heatmap.selected_features(feat_list=superclass_list_pos, sel_feat=sel_pls_pos$`_selected_variables_`, filename="plots/pos_ms2_superclasses_select_pls.pdf", main="PLS", plot_width=6, plot_height=5, cex_col=0.5, cex_row=0.4)
```
